Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations

### 2017-08-18 TexSubmission/ParisEM.tex

The exact computation of these expectations is usually not possible in the case of partially observed diffusions. In this paper, we propose to use Sequential Monte Carlo (SMC) methods to approximate smoothing distributions with random particles associated with importance weights.
 \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with resampling steps to duplicate or discard particles according to their importance weights. Unfortunately, these methods cannot be applied directly to partially observed stochastic differential equations since some elementary quantities, such as transition densities of the hidden states, are not available explicitly. Discretization procedures may be used to approximate transition densities, for instance the Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or Gaussian based approximations using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}. Other approaches based on Hermite polynomials expansion were also introduced by \cite{ait-sahalia:1999,ait-sahalia:2002,ait-sahalia:2008} and extended in several directions recently, see \cite{li:2013} and all the references on the approximation of transition densities therein. However, even the most recent discretization based approximations of the transition densities induce a systematic bias of particle based approximations of posterior distributions, see for instance \cite{delmoral:jacod:protter:2001}. To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE).  In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities, i.e. no discretization scheme is used.

 The only solution to solve the smoothing problem for partially observed SDE using SMC methods without any discretization procedure has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}. 
 Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias that does not vanish as the number of particles grows to infinity.
In the case of discrete time state space models, approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}. 
Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. Then, a backward pass is performed to compute new weights (FFBS) or sample trajectories (FFBSi) in order to approximate the smoothing distributions. Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e.\ using the observations as they are received) smoothed expectations of additive functionals. 
Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.  One of the best features of PaRIS algorithm is that it may be implemented online, using the observations $(Y_k)_{k\ge 1}$ as they are received, without any increasing storage requirements. The total number $n$ of observations is fixed in this paper to describe the algorithm but it could process any additional observation straightforwardly.

@@

PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions. It combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm. It does not produce an approximation of the smoothing distributions but of the smoothed expectation of a fixed additive functional and thus  may be used to approximate \eqref{def:addfunc}. Its crucial property is that it does not require a backward pass, the smoothed expectation is computed on-the-fly with the particle filter and no storage of the particles or weights is needed.

@@

In general situations, PaRIS algorithm cannot be used for stochastic differential equations as $\qk$ is unknown. Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio of \cite{douc:garivier:moulines:olsson:2011} is not tractable. Following \cite{fearnhead:papaspiliopoulos:roberts:2008,olsson:strojby:2011}, filtering weights can be approximated by replacing $\qk(\xi^{\ell}_{k},\xi_{k+1}^{i})$  by an unbiased  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable in $\mathbb{R}^q$ such that:
\[
\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)> 0~~\text{a.s}\quad\mbox{and}\quad
\mathbb{E}\left[\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)\middle| \mathcal{G}_{k+1}^N\right] = \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})\eqsp,
\]
where, for all $0\le k \le n$, 
\begin{align*}
\mathcal{F}_{k}^N &= \sigma\left\{Y_{0:k};(\xi^{\ell}_{u},\omega^{\ell}_{u},\tau^{\ell}_{u});J_{v}^{\ell,j};~1\le \ell\le N,~0\le u\le k, 1\le j \le \widetilde{N}, 0\le v< k\right\}\eqsp,\\
\mathcal{G}_{k+1}^N &= \mathcal{F}_{k}^N \vee \sigma\left\{Y_{k+1};(\xi^{\ell}_{k+1},\omega^{\ell}_{k+1});~1\le \ell\le N\right\}\eqsp.
\end{align*}
Practical choices for $\zeta_k$ are discussed below, see for instance \eqref{eq:GPE1} which presents the choice made for the implementation of such estimators in our context. In the case where $\qk$ is unknown, the filtering weights in \eqref{eq:importance:weights} then become:
\begin{equation}
\label{eq:random:weight}
\widehat{\omega}^{\ell}_k := \frac{\widehat{q}_{k-1}(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k;\zeta_{k-1})g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_{k-1} (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
\end{equation}
In Algorithm~\ref{alg:Ozaki:PaRIS}, $M$ independent copies $(\zeta^m_{k-1})_{1\le m \le M}$ of $\zeta_{k-1}$ are sampled and the empirical mean of the associated estimates of the transition density are used to compute $\widehat{\omega}^{\ell}_k$ instead of a single realization. Therefore, to obtain a generalized random version of PaRIS algorithm, we only need to be able to sample from the discrete probability distribution $\Lambda_k^N(i,\cdot)$ in the case of SDE based HMM. Consider the following assumption: for all  $0\leq k\leq n-1$, there exists a random variable $\hat{\sigma}^k_+$ measurable with respect to $\mathcal{G}_{k+1}^N$ such that,
\begin{assumpt}
\label{assumpt:global}
\mathrm{sup}_{x,y,\zeta}\;\widehat{q}_k(x,y;\zeta)\leq \hat{\sigma}^k_+\eqsp.
\end{assumpt}
\begin{lemma}
\label{lem:AR:unbiased}
Assume that $A_1$ holds for some $0\le k\le n-1$.  For all $1\le i \le N$, define the random variable $J_k^i$ as follows:
\begin{algorithmic}
\REPEAT
\STATE Sample independently $\zeta$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
\UNTIL{$U \leq \widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta)/\hat{\sigma}^k_+$.}
\STATE Set $J_k^i = J$.
%\STATE \textsc{Sampling step:} Sample independently $
%\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
%\IF{ $U \leq \widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)/\hat{\sigma}_+$,}
%\STATE Set $J_k^{i,\ell} = J$.
%\ELSE 
%\STATE Return to \textbf{\sc Sampling Step}
%\ENDIF
\end{algorithmic}
Then, the conditional probability distribution given $\mathcal{G}_{k+1}^N$ of $J_{k}^{i}$ is $\Lambda_{k}^N(i,\cdot)$. 
\end{lemma}
\begin{proof}
See Appendix \ref{sec:append:proofs}.
\end{proof}
Note that Lemma~\ref{lem:AR:unbiased} still holds if assumption \eqref{assumpt:global} is relaxed and replaced by one of the two following assumptions:
\begin{assumpt}\label{assumpt:ysupport}
\mathrm{sup}_{j,y,\zeta}\;\widehat{q}_k(\xi^j_k,y,\zeta)\leq \hat{\sigma}^k_+\;.
\end{assumpt}
\vspace{-\baselineskip}
\begin{assumpt}\label{assumpt:local}
\mathrm{sup}_{i,j,\zeta}\;\widehat{q}_k(\xi^j_k,\xi^j_{k+1},\zeta)\leq \hat{\sigma}^k_+\;.
\end{assumpt}
It is worth noting that under assumptions \eqref{assumpt:global} or \eqref{assumpt:ysupport}, the linear complexity property of PaRIS algorithm still holds, whereas if only assumption \eqref{assumpt:local} holds, the algorithm has a quadratic complexity.

@@

\begin{algorithm}
\caption{GRand PaRIS algorithm}
\begin{algorithmic}
\FORALL{$i \in 1,\dots, N$}
\STATE Sample $\xi_0^i \sim\eta_0$, $\tau_0^i = 0$  and  $\widehat \omega_0^i = g_0(\xi_0^i)\chi_0(\xi_0^i)/\eta_0(\xi_0^i)$.
\ENDFOR
\FOR{$k \in 0,\dots, n-1$}
\FORALL{$i \in 1,\dots, N$}
\STATE Set $\tau_{k+1}^i=0$;
\STATE Sample $I_{k+1}^{i}$ in $\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1\vartheta_{k+1}(\xi_{k}^1),\dots,\widehat{\omega}_{k}^N\vartheta_{k+1}(\xi_{k}^N)\}$.
\STATE Sample $\xi_{k+1}^{i} \sim p_k(\xi_{k}^{I_{k+1}^{i}},\cdot)$.
\STATE For all $1\le m\le M$, sample independently $\zeta_k^m=(\kappa_m,\mw_m, (U_j^m)_{1\leq j\leq \kappa_m})$ with $\kappa_m\sim \mu$, $\mw_m\sim \mathbb{W}_k^{X_{k+1}}$ and $(U_j^m)_{1\leq j\leq \kappa_m}\sim \mathcal{U}[0,\Delta_k]^{\otimes \kappa_m}$.
\STATE Compute $\widehat{\omega}^{i}_{k+1}$ using equation \eqref{eq:random:weight}.
\FORALL{$\ell \in 1,\dots, \widetilde N$} 
\STATE Sample $J_k^{i,\ell}$ as in Lemma~\ref{lem:AR:unbiased}.
\STATE Update $\tau_{k+1}^i = \tau_{k+1}^i + (\tau^{J_k^{i,\ell}}_{k} + h_k(\xi^{J_k^{i,\ell}}_{k},\xi^i_{k+1}))/\tilde{N}$.
%\STATE \textbf{\sc Sampling Step} \\
%Sample independently $\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to  $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
%\IF{ $$U \leq \frac{\widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)}{\hat{\sigma}^k_+},$$}
%\STATE Set $J_k^{i,\ell} = j$.
%\STATE Update $\tau_{k+1}^i = \tau_{k+1}^i + \left(\tau^{J_k^{i,\ell}}_{k} + h_k(\xi^{J_k^{i,\ell}}_{k},\xi^i_{k+1})\right)/\tilde{N}$;
%\ELSE 
%\STATE Return to \textbf{\sc Sampling Step}
%\ENDIF
\ENDFOR
\ENDFOR
\ENDFOR
\end{algorithmic}
\label{alg:Ozaki:PaRIS}
\end{algorithm}

@@

In order to evaluate the performance of the proposed algorithm, the following strategy has been chosen. We compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method of \cite{olsson:strojby:2011}, for different values of the lag (namely, 1,2,5,10,50). The particle approximation of $\mathcal{Q}(\theta,\theta)$ for each model  is computed using each algorithm, see Figure~\ref{fig:res:SINE:EM} for the SINE model (and respectively Figure~\ref{fig:res:LG:EM} for the log-growth model). This estimation is performed 200 times to obtain the estimates $\hQ_1,\dots,\hQ_{200}$, using $\tilde{N}=2$ particles for PaRIS algorithm, and $M=30$ replications for the Monte Carlo approximation $\widehat q_k$ of each $q_k$.  Moreover, the E step requires the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.  $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, and denoted by $\hQ_\star$. Figure \ref{fig:res:SINE:EM} (resp. \ref{fig:res:LG:EM}) shows this estimate for an example on one simulated data set. The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations require similar computational times.

\subsection*{The SINE model} 
The performance of the GRand PaRIS algorithm are first highlighted using the SINE model, where $(X_t)_{t\geq 0}$ is supposed to be the solution to: 
\begin{equation}
\rmd X_t = \sin \left(X_t-\theta\right)\rmd t + \rmd W_t,~~X_0=x_0\eqsp. \label{eq:Lamp:SINE}
\end{equation}
This simple model has no explicit transition density, however GPE estimators may be computed by simulating Brownian bridges.
The process solution to \eqref{eq:Lamp:SINE} is observed regularly at times $t_0=0,\ldots,t_{100}=50$ through the observation process $(Y_k)_{0\leq k \leq 100}$:
\begin{equation}
Y_k = X_k + \varepsilon_k\label{eq:obs:SINE}\eqsp,
\end{equation}
where the $(\varepsilon_k)_{0\leq k \leq 100}$ are i.i.d. $\mathcal{N}(0,1)$.
In the example displayed on Figure \ref{fig:res:SINE:obs}, we set $\theta=0$.
In that case, the function $\rho_{\Delta_k}$ defined in \eqref{eq:rho:func} can be upper bounded either on $(x,y)$ or only on $y$, the GRand PaRIS algorithm has therefore a linear complexity.

This same experiment was reproduced on 100 different simulated data sets.  For each simulation $s$, the empirical absolute relative bias $\mathsf{arb}_s$ and the empirical absolute coefficient of variation $\mathsf{acv}_s$ are computed as
\begin{align}
\mathsf{arb}_s &= \frac{\vert m(\hQ^s)-\hQ^s_\star\vert }{\vert \hQ^s_\star\vert }\label{eq:emp:bias}\\
\mathsf{acv}_s&=\frac{\sigma(\hQ^s)}{\vert m(\hQ^s)\vert }\label{eq:emp:cv}
\end{align}
where $m(\hQ^s)$ and $\sigma(\hQ^s)$ are the empirical mean and standard deviation of the sample $Q_1^s,\dots,Q_{200}^s$. For each estimation method, the resulting distributions of $\mathsf{arb}_1,\dots,\mathsf{arb}_{100}$ and $\mathsf{acv}_1,\dots,\mathsf{acv}_{100}$  are shown on Figures \ref{fig:mult:SINE:b} and  \ref{fig:mult:SINE:cv}.

@@

\begin{figure}[p]
\centering
\includegraphics[scale=0.5]{res_Estep_SINE}
\caption{{\em SINE model - EM intermediate quantity}. Estimation of the EM intermediate quantity $\mathcal{Q}(\theta,\theta)$  using the fixed lag (FL) technique for 5 different lags, and the GRand PaRIS algorithm using 200 replicates. The whiskers represent the extent of the 95\% central values. The dot represents the empirical mean over the 200 replicates. The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=5000$ particles.}
\label{fig:res:SINE:EM}
\end{figure}

@@

Note that in this case the Lamperti transform to obtain a diffusion with a unitary diffusion term depends on $\sigma$. The process $(X_t)_{t\ge 0}$ is a function of $\sigma$ and is not directly observed if $\sigma$ is unknown, which prevents a direct use of an EM algorithm to estimate $\sigma$. Following \cite[Section~8.2]{beskos:papaspiliopoulos:roberts:fearnhead:2006}, this may be overcome with a two-step transformation of the process $(Z_t)_{t\ge 0}$. After the Lamperti transform, the process is transformed into a diffusion bridge  is used to obtai

\begin{figure}[p]
\centering
\includegraphics[scale=0.5]{obs_LG}
\caption{{\em Log-growth model - observations}. Process $X$ solution to the SDE (balls) and observations $Y$ (circles) at times $t_0=0,\dots,t_{100}=50$.}
\label{fig:res:LG:obs}
\end{figure}

>>>>

The exact computation of these expectations is usually not possible in the case of partially observed diffusions. 
In this paper, we propose to use Sequential Monte Carlo (SMC) methods to approximate smoothing distributions with random particles associated with importance weights.
 \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with resampling steps to duplicate or discard particles according to their importance weights.

Unfortunately, these methods cannot be applied directly to partially observed stochastic differential equations since some elementary quantities, such as transition densities of the hidden states, are not available explicitly.
  Discretization procedures may be used to approximate transition densities, for instance the Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or Gaussian based approximations using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}.
   Other approaches based on Hermite polynomials expansion were also introduced by \cite{ait-sahalia:1999,ait-sahalia:2002,ait-sahalia:2008} and extended in several directions recently, see \cite{li:2013} and all the references on the approximation of transition densities therein.
    However, even the most recent discretization based approximations of the transition densities induce a systematic bias of particle based approximations of posterior distributions, see for instance \cite{delmoral:jacod:protter:2001}.
     To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE). 
     In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities, i.e. no discretization scheme is used.

 The only solution to solve the smoothing problem for partially observed SDE using SMC methods without any discretization procedure has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}. 
 Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias, due to the fixed lag approximation, that does not vanish as the number of particles grows to infinity.
In the case of discrete time state space models, approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}. 
Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. 
Then, a backward pass is performed to compute new weights (FFBS) or sample trajectories (FFBSi) in order to approximate the smoothing distributions. Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e.\ using the observations as they are received) smoothed expectations of additive functionals. 
Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.
 One of the best features of PaRIS algorithm is that it may be implemented online, using the observations $(Y_k)_{k\ge 1}$ as they are received, without any increasing storage requirements.
 The total number $n$ of observations is fixed in this paper to describe the algorithm but it could process any additional observation straightforwardly.

@@

The PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions.
 It combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm.
  It does not produce an approximation of the smoothing distributions but of the smoothed expectation of a fixed additive functional and thus  may be used to approximate \eqref{def:addfunc}.
   Its crucial property is that it does not require a backward pass, the smoothed expectation is computed on-the-fly with the particle filter and no storage of the particles or weights is needed.

@@

In general situations, PaRIS algorithm cannot be used for stochastic differential equations as $\qk$ is unknown. 
Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio of \cite{douc:garivier:moulines:olsson:2011} is not tractable. 
Following \cite{fearnhead:papaspiliopoulos:roberts:2008,olsson:strojby:2011}, filtering weights can be approximated by replacing $\qk(\xi^{\ell}_{k},\xi_{k+1}^{i})$  by an unbiased  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable in $\mathbb{R}^q$ such that:
\[
\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)> 0~~\text{a.s}\quad\mbox{and}\quad
\mathbb{E}\left[\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)\middle| \mathcal{G}_{k+1}^N\right] = \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})\eqsp,
\]
where, for all $0\le k \le n$, 
\begin{align*}
\mathcal{F}_{k}^N &= \sigma\left\{Y_{0:k};(\xi^{\ell}_{u},\omega^{\ell}_{u},\tau^{\ell}_{u});J_{v}^{\ell,j};~1\le \ell\le N,~0\le u\le k, 1\le j \le \widetilde{N}, 0\le v< k\right\}\eqsp,\\
\mathcal{G}_{k+1}^N &= \mathcal{F}_{k}^N \vee \sigma\left\{Y_{k+1};(\xi^{\ell}_{k+1},\omega^{\ell}_{k+1});~1\le \ell\le N\right\}\eqsp.
\end{align*}
Practical choices for $\zeta_k$ are discussed below, see for instance \eqref{eq:GPE1} which presents the choice made for the implementation of such estimators in our context. 
In the case where $\qk$ is unknown, the filtering weights in \eqref{eq:importance:weights} then become:
\begin{equation}
\label{eq:random:weight}
\widehat{\omega}^{\ell}_k := \frac{\widehat{q}_{k-1}(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k;\zeta_{k-1})g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_{k-1} (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
\end{equation}
In Algorithm~\ref{alg:Ozaki:PaRIS}, $M$ independent copies $(\zeta^m_{k-1})_{1\le m \le M}$ of $\zeta_{k-1}$ are sampled and the empirical mean of the associated estimates of the transition density are used to compute $\widehat{\omega}^{\ell}_k$ instead of a single realization. 
Therefore, to obtain a generalized random version of PaRIS algorithm, we only need to be able to sample from the discrete probability distribution $\Lambda_k^N(i,\cdot)$ in the case of SDE based HMM.

Consider the following assumption: for all  $0\leq k\leq n-1$, there exists a random variable $\hat{\sigma}^k_+$ measurable with respect to $\mathcal{G}_{k+1}^N$ such that,
\begin{assumpt}
\label{assumpt:global}
\mathrm{sup}_{x,y,\zeta}\;\widehat{q}_k(x,y;\zeta)\leq \hat{\sigma}^k_+\eqsp.
\end{assumpt}
\begin{lemma}
\label{lem:AR:unbiased}
Assume that \eqref{assumpt:global} holds for some $0\le k\le n-1$.  For all $1\le i \le N$, define the random variable $J_k^i$ as follows:
\begin{algorithmic}
\REPEAT
\STATE Sample independently $\zeta$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
\UNTIL{$U \leq \widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta)/\hat{\sigma}^k_+$.}
\STATE Set $J_k^i = J$.
%\STATE \textsc{Sampling step:} Sample independently $
%\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
%\IF{ $U \leq \widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)/\hat{\sigma}_+$,}
%\STATE Set $J_k^{i,\ell} = J$.
%\ELSE 
%\STATE Return to \textbf{\sc Sampling Step}
%\ENDIF
\end{algorithmic}
Then, the conditional probability distribution given $\mathcal{G}_{k+1}^N$ of $J_{k}^{i}$ is $\Lambda_{k}^N(i,\cdot)$. 
\end{lemma}
\begin{proof}
See Appendix \ref{sec:append:proofs}.
\end{proof}
Note that Lemma~\ref{lem:AR:unbiased} still holds if assumption \eqref{assumpt:global} is relaxed and replaced by one of the two following assumptions:
\begin{assumpt}\label{assumpt:ysupport}
\mathrm{sup}_{j,y,\zeta}\;\widehat{q}_k(\xi^j_k,y,\zeta)\leq \hat{\sigma}^k_+\;.
\end{assumpt}
\vspace{-\baselineskip}
\begin{assumpt}\label{assumpt:local}
\mathrm{sup}_{i,j,\zeta}\;\widehat{q}_k(\xi^j_k,\xi^i_{k+1},\zeta)\leq \hat{\sigma}^k_+\;.
\end{assumpt}
It is worth noting that under assumptions \eqref{assumpt:global} or \eqref{assumpt:ysupport}, the linear complexity property of PaRIS algorithm still holds, whereas if only assumption \eqref{assumpt:local} holds, the algorithm has a quadratic complexity.

@@

\begin{algorithm}
\caption{GRand PaRIS algorithm}
\begin{algorithmic}
\FORALL{$i \in 1,\dots, N$}
\STATE Sample $\xi_0^i \sim\eta_0$, $\tau_0^i = 0$  and  $\widehat \omega_0^i = g_0(\xi_0^i)\chi_0(\xi_0^i)/\eta_0(\xi_0^i)$.
\ENDFOR
\FOR{$k \in 0,\dots, n-1$}
\FORALL{$i \in 1,\dots, N$}
\STATE Set $\tau_{k+1}^i=0$;
\STATE Sample $I_{k+1}^{i}$ in $\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1\vartheta_{k+1}(\xi_{k}^1),\dots,\widehat{\omega}_{k}^N\vartheta_{k+1}(\xi_{k}^N)\}$.
\STATE Sample $\xi_{k+1}^{i} \sim p_k(\xi_{k}^{I_{k+1}^{i}},\cdot)$.
\STATE For all $1\le m\le M$, sample independently $\zeta_k^m=(\kappa_m,\mw_m, (U_j^m)_{1\leq j\leq \kappa_m})$ with $\kappa_m\sim \mu$, $\mw_m\sim \mathbb{W}_k^{X_{k+1}}$ and $(U_j^m)_{1\leq j\leq \kappa_m}\sim \mathcal{U}[0,\Delta_k]^{\otimes \kappa_m}$.
\STATE Compute $\widehat{\omega}^{i}_{k+1}$ using equation \eqref{eq:random:weight}.
\FORALL{$\ell \in 1,\dots, \widetilde N$} 
\STATE Sample $J_k^{i,\ell}$ as in Lemma~\ref{lem:AR:unbiased}.
\STATE Update $\tau_{k+1}^i = \tau_{k+1}^i + (\tau^{J_k^{i,\ell}}_{k} + h_k(\xi^{J_k^{i,\ell}}_{k},\xi^i_{k+1}))/\tilde{N}$.
\ENDFOR
\ENDFOR
\ENDFOR
\end{algorithmic}
\label{alg:Ozaki:PaRIS}
\end{algorithm}

@@

In order to evaluate the performance of the proposed algorithm, the following strategy has been chosen. 
We compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method of \cite{olsson:strojby:2011}, for different values of the lag (namely, 1,2,5,10,50). 
The particle approximation of $\mathcal{Q}(\theta,\theta)$ for each model  is computed using each algorithm, see Figure~\ref{fig:res:SINE:EM} for the SINE model (and respectively Figure~\ref{fig:res:LG:EM} for the log-growth model). 
This estimation is performed 200 times to obtain the estimates $\hQ_1,\dots,\hQ_{200}$, using $\tilde{N}=2$ particles for PaRIS algorithm, and $M=30$ replications for the Monte Carlo approximation $\widehat q_k$ of each $q_k$.  
Moreover, the E step requires the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.  $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
In order to obtain a reference value for our study, the intermediate quantity of the EM algorithm is also estimated 30 times using the GRand PaRIS algorithm with $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, and denoted by $\hQ_\star$. 
Figure \ref{fig:res:SINE:EM} (resp. \ref{fig:res:LG:EM}) shows this estimate for an example on one simulated data set. 
The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations require similar computational times.

\subsection*{The SINE model} 
The performance of the GRand PaRIS algorithm are first highlighted using the SINE model, where $(X_t)_{t\geq 0}$ is supposed to be the solution to: 
\begin{equation}
\rmd X_t = \sin \left(X_t-\mu\right)\rmd t + \rmd W_t,~~X_0=x_0\eqsp. \label{eq:Lamp:SINE}
\end{equation}
This simple model has no explicit transition density, however GPE estimators may be computed by simulating Brownian bridges.
The process solution to \eqref{eq:Lamp:SINE} is observed regularly at times $t_0=0,\ldots,t_{100}=50$ through the observation process $(Y_k)_{0\leq k \leq 100}$:
\begin{equation*}
Y_k = X_k + \varepsilon_k\eqsp,
\end{equation*}
where the $(\varepsilon_k)_{0\leq k \leq 100}$ are i.i.d. $\mathcal{N}(0, \sigma^2_\text{obs})$, the resulting set of model parameters is $\theta = (\mu, \sigma_\text{obs})$.
In the example displayed on Figure \ref{fig:res:SINE:obs}, we set $\mu=0$ an d $ \sigma_\text{obs} = 1$.
In that case, the function $\rho_{\Delta_k}$ defined in \eqref{eq:rho:func} can be upper bounded either on $(x,y)$ or only on $y$, the GRand PaRIS algorithm has therefore a linear complexity.

This same experiment was reproduced on 100 different simulated data sets. 
For each simulation $s$, the empirical absolute relative bias $\mathsf{arb}_s$ and the empirical absolute coefficient of variation $\mathsf{acv}_s$ are computed as
\begin{align*}
\mathsf{arb}_s &= \frac{\vert m(\hQ^s)-\hQ^s_\star\vert }{\vert \hQ^s_\star\vert }\\
\mathsf{acv}_s&=\frac{\sigma(\hQ^s)}{\vert m(\hQ^s)\vert }\eqsp,
\end{align*}
where $m(\hQ^s)$ and $\sigma(\hQ^s)$ are the empirical mean and standard deviation of the sample $Q_1^s,\dots,Q_{200}^s$. For each estimation method, the resulting distributions of $\mathsf{arb}_1,\dots,\mathsf{arb}_{100}$ and $\mathsf{acv}_1,\dots,\mathsf{acv}_{100}$  are shown on Figures \ref{fig:mult:SINE:b} and  \ref{fig:mult:SINE:cv}.

@@

\begin{figure}[p]
\centering
\includegraphics[scale=0.5]{res_Estep_SINE}
\caption{{\em SINE model - EM intermediate quantity}. Estimation of the EM intermediate quantity $\mathcal{Q}(\theta,\theta)$  using the fixed lag (FL) technique for 5 different lags, and the GRand PaRIS algorithm using 200 replicates. 
The whiskers represent the extent of the 95\% central values. 
The dot represents the empirical mean over the 200 replicates. 
The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N = 5000$ particles.}
\label{fig:res:SINE:EM}
\end{figure}

@@

Note that in this case the Lamperti transform to obtain a diffusion with a unitary diffusion term depends on $\sigma$. 
The process $(X_t)_{t\ge 0}$ is a function of $\sigma$ and is not directly observed if $\sigma$ is unknown, which prevents a direct use of an EM algorithm to estimate $\sigma$. 
Following \cite[Section~8.2]{beskos:papaspiliopoulos:roberts:fearnhead:2006}, this may be overcome with a two-step transformation of the process $(Z_t)_{t\ge 0}$.
\begin{figure}[p]
\centering
\includegraphics[scale=0.5]{obs_LG}
\caption{{\em Log-growth model - observations}.
 Process $X$ solution to the SDE (balls) and observations $Y$ (circles) at times $t_0=0,\dots,t_{100}=50$.}
\label{fig:res:LG:obs}
\end{figure}
